Comparing vector-based and ACT-R memory models of tag retrieval:\protect\\User-customized hashtag/tag prediction on Twitter and StackOverflow

### 2014-01-07 writeups/proposal/proposal.tex
merge
Former-commit-id: 83c281e059e8ae39158f65201aed9e4ab8b2e054 [formerly 0bfc9115dd3c3a151aa40aaa7afcc129920e6c83]
Former-commit-id: b396325413ad09575644838d019f84cc81e87c04

@@

ACT-R's declarative memory and vector-based memory systems are substantially different in their mathematical structure.
Nonetheless, both can successfully model classic behavioral patterns found in word pairing experiments (e.g., \cite{Rutledge2008} modeling the ``fan effect'').
I will extend the work done by \citeauthor{Rutledge2008} by formally comparing these two models on large-scale hashtag prediction tasks.
I will also replace the vector-based model used by \citeauthor{Rutledge2008} with an improved and simplified random-permutation model \parencite{Sahlgren2008}.
The primary objective is to compare the random-permutation vector-based model to ACT-R's declarative memory theory.

@@

A user's prior hashtag use will be an essential model component for domains such as Twitter, where the number of possible hashtags is practically infinite.
In these domains, the model can utilize a user's hashtag history as a way to prune the infinite space of possible hashtags, and generate a much smaller and more manageable set for prediction.
However, not only should the model properly take into account a user's prior hashtag use, but it should also generate new hashtag predictions when a user is most likely choosing a hashtag they have never used before.
Calibrating the models to properly balance this exploitation of prior hashtag use versus exploration of new hashtags will be a primary research question.
The end goal is to have a model that can properly balance the user's tag history with the contextual cues in the post in order to generate an accurate prediction.

>>>>

The ACT-R cognitive architecture \parencite{Anderson2004} contains one of the leading theories of human declarative memory.
The declarative memory component in ACT-R is a Bayesian statistical model.
This is the same type of co-occurrence Naive Bayes statistical model that is prominent in the more general machine-learning field.
When these Bayesian models are used for machine learning, it is common to use a large number of observations to construct the co-occurrence matrix that is the foundation for the model.
The primary reason for this is that the stability of statistical models increase as sample size increases.

However it is much less common in the cognitive modeling community to use a large number of observations to build up declarative memory when creating ACT-R models.
If a cognitive modeler is using the declarative memory component of ACT-R,
it is common to either construct the co-occurrences by hand, or use a small training set of data to initially build up the declarative memory system.
This approach can and has worked well.
Essentially the modeler is only including the observations in declarative memory that are relevant to the specific task that they are studying.

This approach is problematic for two main reasons.
First, it means that the declarative memory system for every developed model must be hand crafted and customized for that model, which takes enormous time and effort.
Second, it means that modelers are continuing to use small sample sizes for the declarative memory components in their models.
This has important implications for understanding declarative memory:
It means that we do yet fully understand how well the ACT-R declarative memory theory scales when the number of chunks begins to approach the number of chunks that are actually stored in human declarative memory.

There is, however, an opportunity to change this approach.
With the massive growth of human-created content on social media sites (e.g., Twitter, StackOverflow, Wikipedia), and public access to many of those databases,
it is now possible to evaluate psychological theories of declarative memory on large-scale real-world tasks where the theory can be tested on hundreds of millions to billions of data points.
This will provide a way to test if these models scale, and evaluate the psychological validity of these models as the number of observations used increases by orders of magnitude.

There have already been a few researchers that have used these large databases to evaluate how well declarative memory models scale,
for both ACT-R Bayesian models \parencites{Fu2007,Pirolli2003,Stanley2013,Douglass2010} and vector-based memory models \parencites{Jones2007,Rutledge2008,Recchia2010,Sahlgren2008}.

The work for this proposal will expand on the knowledge gained from previous work that has tested retrieval models on large-scale datasets.
It will focus specifically on how two state-of-the-art declarative memory retrieval models (ACT-R Bayesian and random permutations) can be improved further, by testing these models on two real-world tagging tasks.
One primary objective is to compare the random-permutation vector-based model to ACT-R's declarative memory theory when the datasets used for each model contain hundreds of millions to billions of data points.
Another objective is to test how much improvement is gained by incorporating the strengths of each model into the other (i.e., modifying model components).
For example, it is currently unexplored how a vector-based model can be modified to retrieve results that are based not only on context, but also on the prior odds that a particular item in memory will be needed again.
Since large datasets will be used, a rich volume of data will be available to test the models.
This enables constraints and assumptions for each theory to be rapidly tested and new model components to be hypothesized and quickly evaluated.

\subsection{Motivation for Task}

@@

ACT-R's declarative memory and vector-based memory systems are substantially different in their mathematical structure.
Nonetheless, both can successfully model classic behavioral patterns found in word pairing experiments (e.g., \cite{Rutledge2008} modeling the ``fan effect'').
I will extend the work done by \citeauthor{Rutledge2008} by formally comparing these two models on large-scale hashtag prediction tasks.
I will also replace the vector-based model used by \citeauthor{Rutledge2008} with an improved and simplified random-permutation model \parencite{Sahlgren2008}.

@@

A user's prior hashtag use will be an essential model component for domains such as Twitter, where the number of possible hashtags is practically infinite.
In these domains, the model can utilize a user's hashtag history as a way to prune the infinite space of possible hashtags, and generate a much smaller and more manageable set for prediction.
However, not only should the model properly take into account a user's prior hashtag use, but it should also generate new hashtag predictions when a user is most likely choosing a hashtag they have never used before.
The end goal is to have a model that can properly balance the user's tag history with the contextual cues in the post in order to generate an accurate prediction.
